---
title: "DATA420 Monte Carlo Simulation"
author: "Jason Pemberton"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

Basic understanding of a Monte Carlo Simulation:
We will look at an example where we simulate throwing 1000 darts at a dart board (circle).

The X-Coordinates of the circle range from -1 to 1 (along the x-axis), while the Y-Coordinates range from -1 to 1 along the y-axis. We will use the runif function to randomly select a set of coordinates along each axis, wrapping this function in a for loop to allo us to run 1000 simulated "dart throws".

For each "dart throw" we will check whether the outcome lies within the circle by using the formula:

X^2 + y^2 <= 1

We will use the results of our model to see how well it estimates the value of pi. Recall that the area of a circle is (pi * radius^2). In our example we can calculate pi by using the formula: pi = 4 * number of darts in circle / number of darts thrown

```{r Monte Carlo Simulation}
# Number of simulations to run
num_darts <- 10000

# Start our counter at zero
num_darts_in_circle <- 0

for (i in 1:num_darts) {
  x <- runif(n=1, min=-1, max=1)
  y <- runif(n=1, min=-1, max=1)
  if (x^2 + y^2 <= 1) {
    num_darts_in_circle = num_darts_in_circle + 1
  }
}

paste("Model accuracy: ", num_darts_in_circle/num_darts*100, "%")

# pi = 3.14159265359
paste("Model estimate of pi: ", 4 * num_darts_in_circle / num_darts)
```

We can achieve a better estimate of pi by increasing the number of darts throw - however this will significantly increase run time. We can optimize this. R can be slow when running iterating through loops but can be quite fast when working with vector and matrix operations.

We will remove the for and if loops and replace them with vectors.
```{r Fast Monte Carlo}
# Number of simulations to run
num_darts <- 10000

# Start our counter at zero
num_darts_in_circle <- 0

# Vectors of length num_darts
x <- runif(n=num_darts, min=-1, max=1)
y <- runif(n=num_darts, min=-1, max=1)

# Also a vector of length num_darts
sum_squares <- x^2 + y^2
indexes_darts_in_circle <- which(sum_squares <= 1)

num_darts_in_circle2 <- length(indexes_darts_in_circle)

paste("Model accuracy: ", num_darts_in_circle2/num_darts*100, "%")

# pi = 3.14159265359
paste("Model estimate of pi: ", 4 * num_darts_in_circle2 / num_darts)
```

Depending on your computer performance, model run time can differ significantly (in my case 1,000,000 simulated dart throws took 8 seconds using for and if loops, and less than a second with vector calculation)

```{r Visualize the Results}

# Create a data frame with the generated points
df <- data.frame(x, y, in_circle = sum_squares <= 1)

# Create a scatter plot of the points, color them based on whether they are inside the circle or not
plot <- ggplot(df, aes(x, y, color = in_circle)) +
  geom_point() +
  theme_minimal()

# Create a data frame for circle points
circle_df <- data.frame(
  x = cos(seq(0, 2 * pi, length.out = 100)),
  y = sin(seq(0, 2 * pi, length.out = 100))
)

# Add a red circle to the plot
plot <- plot +
  geom_point(data = data.frame(x = 0, y = 0), aes(x, y), color = "red", size = 2) +
  geom_path(data = circle_df, aes(x, y), color = "red")+
  labs(title=paste("Model accuracy: ", num_darts_in_circle2/num_darts*100, "%"))

# Set the aspect ratio to 1
plot <- plot + coord_fixed(ratio = 1)

# Display the plot
print(plot)

```

