---
title: "DATA420 Time Series Forecasting"
author: "Jason Pemberton"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
library(tidyverse)
library(forecast)
library(plotly)
```

**PART 1**: Loading and displaying time series data

```{r Load Data}
temperature <- read_csv("C:/Users/jason/OneDrive/SAIT/SAIT Instructor/COURSE - NEW_SAIT_DATA420/data/temperature.csv")
head(temperature,5)
```

Create a new calculated column called Avg_Temperature using the rowMeans function (don't forget to exclude the first column "Year")

```{r Calculated Column - Average Temp}
temperature$Avg_Temp <- rowMeans(temperature[2:13])
head(temperature,5)
```

```{r Plotting time series data}
data <- data.frame(Year = temperature$year, Temperature = temperature$Avg_Temp)

data %>% 
  ggplot(aes(Year, Temperature))+
  geom_line()+
  geom_point()+
  labs(title="Temperature Trend 2000-2017",
       x = "Year",
       y="Average Temperature")
```

**PART 2**: Forecasting time series data using ARIMA model ARIMA, which stands for Autoregressive Integrated Moving Average, is a widely used statistical method for time series forecasting. It is a type of time series analysis and forecasting model that combines autoregression, differencing, and moving averages.

**Auto Regressive (AR)**: This part of the model uses the relationship between an observation and several lagged observations (previous time steps). It suggests that the current value of a time series is linearly dependent on its past values.

**Integrated (I)**: This component involves differencing the time series to make it stationary. Stationarity is a key assumption in many time series models, including ARIMA. Differencing involves subtracting the previous observation from the current observation, and it helps to remove trends or seasonality.

**Moving Average (MA)**: This part of the model uses the relationship between an observation and a residual error from a moving average model applied to lagged observations. It represents the moving average of past errors.

The ARIMA model takes three parameters ARIMA(p, d, q), where: - p: The order of the autoregressive part. - d: The degree of differencing. - q: The order of the moving average part.

The ARIMA model is suitable for time series data with a clear trend and seasonality. It is used in various fields, such as finance, economics, and environmental science, for forecasting future values based on historical data.

In this example we will use auto.arima. It automatically determines the optimal parameters for an ARIMA model, such as the order of differencing, autoregressive (AR) terms, and moving average (MA) terms.

```{r Plotting with Confidence Intervals}
# Fit ARIMA model
arima_model <- auto.arima(data$Temperature)

forecast_ts <- forecast(arima_model, h = 5, level = c(80, 95))  # forecasting 5 months ahead with 80% and 95% intervals
forecast_ts


# Create a data frame for forecast results and intervals
forecast_data <- data.frame(Year = seq(max(data$Year) + 1, max(data$Year) + 5),
                            Forecast = forecast_ts$mean,
                            CI_80_lower = forecast_ts$lower[, "80%"],
                            CI_80_upper = forecast_ts$upper[, "80%"],
                            CI_95_lower = forecast_ts$lower[, "95%"],
                            CI_95_upper = forecast_ts$upper[, "95%"])

# Add forecast values and intervals to the plot
p <- ggplot() +
  geom_ribbon(data = forecast_data, aes(x = Year, ymin = CI_80_lower, ymax = CI_80_upper), fill = "blue", alpha = 0.2) +
  geom_ribbon(data = forecast_data, aes(x = Year, ymin = CI_95_lower, ymax = CI_95_upper), fill = "blue", alpha = 0.1) +
  geom_line(data = forecast_data, aes(x = Year, y = Forecast, color = "Forecast")) +
  geom_point(data = forecast_data, aes(x = Year, y = Forecast, color = "Forecast")) +
  geom_line(data = data, aes(x = Year, y = Temperature, color = "Observed")) +
  geom_point(data = data, aes(x = Year, y = Temperature, color = "Observed")) +
  labs(title = "Temperature Trend with ARIMA Forecast",
       x = "Year",
       y = "Average Temperature",
       color = "Series")


ggplotly(p)
```








