---
title: "DATA 420 - Time Series Forecasting Example - NDVI"
author: "Jason Pemberton"
date: "2023-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
library(forecast)
library(tidyverse)
library(plotly)
```

This example of using R for time series forcasting focuses on a dataset called NDVI - Landsat Normalized Difference Vegetation Index. This is a set of satellite based observations of vegetation "greenness" that is used to understand vegetation density and assess plant health. Our dataset contains a column NDVI which is a series of measurements gathered over a 12 year time period.

After the data is loaded and prepared for analysis we will create a series of Arima models to forecast NDVI values over time. We will start with a model assuming no seasonality in the data then move up to a more complicated (but also more realistic model) that includes seasonality.

Step 1:
Load the data
Format the date column as.Date
Create a time series dataset
```{r Load and Prepare Data}

# Load the data
data <- read_csv("C:/Users/jason/OneDrive/SAIT/SAIT Instructor/COURSE - NEW_SAIT_DATA420/data/portal_timeseries.csv")
str(data)

# set date format
data$date <- as.Date(data$date, format="%m/%d/%Y")

# create time series object using "ts" function
# NDVI = Normalized Difference Vegetation Index (a dataset to quantify vegetation density and plant health)
NDVI_ts <- ts(data$NDVI, 
              start= c(1992,3),
              end=c(2014,11),
              frequency=12)
```


Step 2: 
Exploratory Data Analysis
Plot the original data
Use the Auto-Correlate function to look for seasonality 

```{r Exploratory Data Analysis}
# plot your data - looking at the NDVI measurements there does not appear to be any sign of a trend or seasonality in the data. 
plot(NDVI_ts)


# Auto-correlate function. Looking for seasonality that the program might detect that we might be missing
# This plot shows some seasonal trends at the 1-2 month level, as well as smaller trends at 6 months, and longer annual 1-2 year cycles.
acf(NDVI_ts)
```

The equation for plotting time series data is:
Yt = c + et

Where:
Yt = value at time t
c = coefficient (average or long term average)
et = error or noise
et = N(mean=0, sd), normally distributed

The ARIMA mode (Auto-Regressive, Integration, Moving Average) takes three input parameters. In our first example we will run the arima model with parameters set to zero
```{r Fitting our Model}

# STEP 4: Choosing and fitting a model (ARIMA) - c(0,0,0) no Auto-Regressive, Integration, or Moving Average applied
avg_model <- arima(NDVI_ts, c(0,0,0))
str(avg_model)

# STEP 5: Forecasting

avg_forecast <- forecast(avg_model)
str(avg_forecast)

avg_forecast$mean
# The values are all the same because there is no time dependence for the NDVI variable. They are just a long term average
```


Our first plot will add a red line to the original data showing the forecast. The line appears flat because there is no relationship between time and NDVI. The model uses the previous data point to forecast the next datapoint, and so the average does not change over time.

The second plot automatically adds the 80 and 90% confidence intervals

The third plto demonstrates a shortcut in R that allows us to quickly produce a ggplot of the data using the function "autoplot"
```{r Visualize the Forecast}
# Visualize our original data
plot(NDVI_ts)
lines(avg_forecast$mean, col="red")

# The forecast package can make these plots for us automatically with confidence intervals as well
plot(avg_forecast)

# Using ggplot instead of base R
autoplot(avg_forecast)
```


Now we will use the auto.arima function to create a model that selects the three input parameters automatically.
The first auto-arima model we set seasonality to False to see how the model performs
The second plot we create a proper seasonal model
This example is more realistic and the forecast is no longer a flat line. Its still not perfect. We can see the effects of decay over time as the moving average decreases with each interval.
```{r Auto Arima}
# Using auto.arima to fit a season forecast
arima_model <- auto.arima(NDVI_ts, seasonal= FALSE)
arima_forecast <- forecast(arima_model)
autoplot(arima_forecast)

# ARIMA forecasts one step at a time into the future, 
# it will begin by looking back at t-1. Over time the influence of the previous value "decays"

seasonal_arima_model <- auto.arima(NDVI_ts)
seasonal_arima_forecast <- forecast(seasonal_arima_model,
                                    h=36,
                                    level=c(80,99))

plot(seasonal_arima_forecast)

```

