---
title: "Penguins"
author: "Jason Pemberton"
date: "2023-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
# Load required libraries
library(tidyverse)
library(rvg)
library(plotly)
```

```{r Import Data, include=FALSE}
#import the data
df <- read.csv("C:/Users/jason/OneDrive/SAIT/SAIT Instructor/COURSE - NEW_SAIT_DATA420/data/penguins.csv")
```

#### **Getting to know the penguins dataframe**

head(), summary(), and glimpse() give us a good look at the shape and basic distribution of our dataframe

Basic R plots such as ***hist()*** (histogram), ***boxplot()***, and ***plot()*** allow us to visualize out basic data distribution

```{r Exploratory Data Analysis, echo=FALSE}

head(df)
summary(df)
glimpse(df)

#Basic R functions for plotting data
hist(df$body_mass_g)
boxplot(df$bill_length_mm)
plot(df$flipper_length_mm, df$bill_length_mm)

plot(df[,c('bill_length_mm', 'bill_depth_mm', 'flipper_length_mm', 'body_mass_g')])
```

**Analysis of penguin species**

ggplot (part of tidyverse) allows us to take our visualizations to the next level.

Scatterplot using two continuous variables

```{r Scatterplot, echo=FALSE, message=FALSE, warning=FALSE}
# Use GGPLOT to create a visual for two continuous variables
df %>% 
  ggplot() + 
  geom_point(mapping=aes(x=bill_length_mm, y=bill_depth_mm, color=species, size=body_mass_g, alpha=0.7))+
  labs(
    title="Penguins Bill Depth vs Bill Length",
    x="Bill Length (mm)",
    y="Bill Depth (mm)")+
  theme(plot.title=element_text(hjust=0.5))

```

Column chart using one categorical variables

```{r Bar Chart, echo=FALSE, message=FALSE, warning=FALSE}
# Use GGPLOT to create a visual for a categorical variable
df[!(is.na(df$sex) | df$sex==""), ] %>% 
  ggplot(mapping=aes(x=species, fill=sex))+
  #geom_bar()+
  geom_bar(position="dodge")+
  labs(
    title="Penguins Species and Sex",
    x="Species",
    y="Count")+
  theme(plot.title=element_text(hjust=0.5))
```

Density plot

```{r Density Plot, echo=FALSE}
df[!(is.na(df$sex) | df$sex==""), ] %>% 
  ggplot(aes(x=body_mass_g, fill=sex))+
  geom_density(alpha=0.5)+
  labs(
    title="Body Mass Distribution",
    x="Body Mass (g)",
    y="Density")+
  theme(plot.title=element_text(hjust=0.5))
```

Box Plot

Solid line = Median

Edges of box = lower quartile (25%) and upper quartile (75%)

points outside whiskers = extremes

```{r Box Plot, echo=FALSE}
df[!(is.na(df$sex) | df$sex==""), ] %>% 
  ggplot(aes(x = species, y = body_mass_g)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "blue")+
    labs(
    title="Body Mass Distribution by Species",
    x="Species",
    y="Body Mass (g)")+
  theme(plot.title=element_text(hjust=0.5))
```
