---
title: "T-Test"
author: "Jason Pemberton"
date: "2025-03-30"
output: html_document
---

```{r Load Libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(patchwork)
library(gapminder)
```

---
Hypothesis Testing:
H0 (null hypothesis) - The mean life expectancy in Afirca is 50 years
H1 (alternative hypothesis) - The mean life expectancy is not 50 years  

Observation: Sample data provides a mean life expectancy for Afirca of 48.9. Is this statistically significant?  

Lets start by assuming the mean life expectancy is actually 50 years.

In most cases, 5% or less is statistically significant (p < 0.05)
```{r t-test, single sample}

gapminder %>% 
  filter(continent == "Africa") %>% 
  select(lifeExp) %>% 
  t.test(mu = 50)

# p-value is 0.002038
# we can reject the null hypothesis
# we accept the alternative hypothesis

my_ttest <- gapminder %>% 
  filter(continent == "Africa") %>% 
  select(lifeExp) %>% 
  t.test(mu = 50)

# what is inside you t-test object?
attributes(my_ttest)

my_ttest$p.value
```

---

``` {r Visualize Africa life expectancy}
# Filter data for Africa
africa_data <- gapminder %>%
  filter(continent == "Africa")

# Calculate mean life expectancy for Africa
mean_lifeExp_africa <- mean(africa_data$lifeExp)

# Create density plot
ggplot(africa_data, aes(x = lifeExp)) +
  geom_density(fill = "red", alpha = 0.2) +
  geom_vline(xintercept = mean_lifeExp_africa, linetype = "dashed", color = "black", size = 1) +
  labs(title = "Density Plot of Life Expectancy in Africa",
       x = "Life Expectancy",
       y = "Density") +
  theme_minimal()
```

---

```{r t-test, two-sided}
# H0: Africa and Europe mean life expectancy are the same
# H1: Africa and Europe mean life expectancy are different

gapminder %>% 
  filter(continent %in% c("Africa", "Europe")) %>%
  t.test(lifeExp ~ continent, data = ., 
         alternative = "two.sided") # data = .by using pipe, it wants to place gapminder as the first argument to t.test

# p-value is < 2.2e-16
# we can reject the null hypothesis
# we accept the alternative hypothesis
```

---

``` {r Visualize the two continents}
# Filter data for Africa and Europe
filtered_data <- gapminder %>%
  filter(continent %in% c("Africa", "Europe"))

# Calculate mean life expectancy for each continent
mean_lifeExp <- filtered_data %>%
  group_by(continent) %>%
  summarise(mean_lifeExp = mean(lifeExp))

# Create density plot
ggplot(filtered_data, aes(x = lifeExp, fill = continent)) +
  geom_density(alpha = 0.5) +
  geom_vline(data = mean_lifeExp, aes(xintercept = mean_lifeExp, color = continent), 
             linetype = "dashed", linewidth=1) +
  labs(title = "Density Plot of Life Expectancy",
       x = "Life Expectancy",
       y = "Density",
       fill = "Continent",
       color = "Mean Life Expectancy") +
  theme_minimal()

```

---

```{r t-test, difference of means}
# H0: Ireland has a mean life expectancy the same as Switzerland
# H1: Ireland mean life expectancy is less

gapminder %>% 
  filter(country %in% c("Ireland", "Switzerland")) %>% 
  t.test(lifeExp ~ country, data = .,
         alternative = "less",
         conf.level = 0.95)

# p-value is 0.05835
# we cannot reject the null hypothesis
# we reject the alternative hypothesis

```

---

``` {r Visualize the two countries}
# Filter data for Ireland and Finland
filtered_data <- gapminder %>%
  filter(country %in% c("Ireland", "Switzerland"))

# Calculate mean life expectancy for each country
mean_lifeExp <- filtered_data %>%
  group_by(country) %>%
  summarise(mean_lifeExp = mean(lifeExp))

# Create density plot with x-axis set from 50 to 90
ggplot(filtered_data, aes(x = lifeExp, fill = country)) +
  geom_density(alpha = 0.5) +
  geom_vline(data = mean_lifeExp, aes(xintercept = mean_lifeExp, color = country), 
             linetype = "dashed", linewidth = 1) +
  scale_x_continuous(limits = c(50, 90)) +  # Manually sets x-axis limits
  labs(title = "Density Plot of Life Expectancy in Ireland & Switzerland",
       x = "Life Expectancy",
       y = "Density",
       fill = "Country",
       color = "Mean Life Expectancy") +
  theme_minimal()
```

---

```{r t-test, paired}
# H0: The mean life expectancy for Africa is the same in 1957 and 2007
# H1: They are different


#gapminder %>% 
#  filter(year %in% c(1957, 2007) &
#           continent == "Africa") %>% 
#  mutate(year = factor(year, levels = c(2007, 1957))) %>% 
#  t.test(lifeExp ~ year, data = .,
#         paired = TRUE)

# this doesn't work

# Filter for African countries in 1957 and 2007
paired_data <- gapminder %>%
  filter(continent == "Africa" & year %in% c(1957, 2007)) %>%
  group_by(country) %>%
  filter(n() == 2) %>%  # Keep only countries with both years
  ungroup() %>%
  arrange(country, year)  # Ensure correct pairing order

# Perform paired t-test
 
t.test(gapminder$lifeExp[gapminder$year == 2007], gapminder$lifeExp[gapminder$year == 1957], paired = TRUE, data = paired_data)

# p-value is < 2.2e-16
# we reject the null hypothesis
# we accept the alternative hypothesis
```

---

```{r}
# Filter data for Africa in 1957 and 2007
africa_data <- gapminder %>%
  filter(continent == "Africa" & year %in% c(1957, 2007)) %>%
  mutate(year = factor(year))  # Convert year to a factor for plotting

# Calculate mean life expectancy for each year
mean_lifeExp <- africa_data %>%
  group_by(year) %>%
  summarise(mean_lifeExp = mean(lifeExp))

# Create density plot
ggplot(africa_data, aes(x = lifeExp, fill = year)) +
  geom_density(alpha = 0.5) +  # Density plot with transparency
  geom_vline(data = mean_lifeExp, aes(xintercept = mean_lifeExp, color = year), 
             linetype = "dashed", linewidth = 1) +  # Add dashed lines for mean values
  labs(title = "Density Plot of Life Expectancy in Africa (1957 vs 2007)",
       x = "Life Expectancy",
       y = "Density",
       fill = "Year",
       color = "Mean Life Expectancy") +
  theme_minimal()
```

---

H0: The average height of male and female Star Wars characters is the same
H1: The average height is not the same

```{r T-Test with Star Wars Database, echo=FALSE, message=FALSE, warning=FALSE}

my_data <- starwars %>% 
  select(sex, height) %>% 
  filter(sex %in% c("male", "female")) %>% 
  drop_na(height)

t.test(height ~ sex, data = my_data)
# The average heights of female and male characters is not the same, but the p-value is > 0.05 (0.2431)
# We cannot reject the null hypothesis. The difference we are seeing could just be by chance

# same result using pipe operator in tidyverse
# starwars %>% 
#   select(sex, height) %>% 
#   filter(sex %in% c("male", "female")) %>% 
#   drop_na(height) %>% 
#   t.test(height ~ sex, data = .)
  

```

