---
title: "ANOVA"
author: "Jason Pemberton"
date: "2025-03-31"
output: html_document
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(patchwork)
library(gapminder)
library(forcats)
```

---

```{r Create Dataset}

gapdata <- gapminder %>% 
  filter(year == 2007 &
           continent %in% c("Americas", "Europe", "Asia")) %>% 
  select(continent, lifeExp)
```

---
H0 (Null Hypothesis): The mean life expectancy of the three continents is the same  

H1: (Alternative Hypothesis): The mean life expectancy is not the same  

```{r Look at distribution of means}

gapdata %>%
  group_by(continent) %>%
  summarise(Mean_life = mean(lifeExp)) %>%
  arrange(desc(Mean_life))
```

A difference in the mean values is observed in the sample data, but is it statistically significant (p < 0.05)?  

```{r Create ANOVA model}

gapdata %>% 
  aov(lifeExp ~ continent, data = .) %>% 
  summary()

aov_model <- gapdata %>% 
    aov(lifeExp ~ continent, data = .)

```

Is this significance being driven by a particular continent?  
Given than p is much less than 0.05 (p = 0.00000342), we reject the null hypothesis. The mean life expectancy is not the same for each continent.  

---

```{r Compare results}

gapdata %>% 
  aov(lifeExp ~ continent, data = .) %>% 
  TukeyHSD()  # Honestly Significant Difference


gapdata %>% 
  aov(lifeExp ~ continent, data = .) %>% 
  TukeyHSD() %>% 
  plot()
```

The difference between Asia and Americas is not statistically significant, but Europe-Americas, and Europe-Asia is.

---

```{r Visualize}

gapdata %>% 
  ggplot(aes(continent, lifeExp, fill=continent)) +
  geom_boxplot(show.legend = TRUE) +
  geom_jitter(show.legend = FALSE, width = 0.25, shape = 21, color = "black") +
  coord_flip()
```

---

```{r}
gapdata %>%
  ggplot(aes(x = lifeExp, fill = continent, color = continent)) +
  geom_density(alpha = 0.4) +
  # Add mean vertical line for each continent
  geom_vline(data = gapdata %>% 
               group_by(continent) %>% 
               summarise(mean_lifeExp = mean(lifeExp)), 
             aes(xintercept = mean_lifeExp, color = continent), 
             linetype = "dashed") +
  facet_wrap(~ continent, scales = "free", ncol = 1) +  # Stack plots vertically
  labs(title = "Density Plots of Life Expectancy by Continent with Mean Lines",
       x = "Life Expectancy",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend for cleaner appearance
```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)


boxplot <- gapdata %>%
  ggplot(aes(continent, lifeExp, fill = continent)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.5) +  # Remove legend
  geom_jitter(show.legend = FALSE, width = 0.25, shape = 21, color = "black") +
  coord_flip() +
  scale_fill_manual(values = c("Europe" = "blue", "Asia" = "green", "Americas" = "red")) +  # Custom color order
  labs(title = "Life Expectancy by Continent",
       x = "Continent",
       y = "Life Expectancy") +
  theme_minimal()

# Density plot with vertical mean lines (on the right)
density_plot <- gapdata %>%
  mutate(continent = factor(continent, levels = c("Americas", "Asia", "Europe"))) %>% 
  ggplot(aes(x = lifeExp, fill = continent, color = continent)) +
  geom_density(alpha = 0.5) +
  geom_vline(data = gapdata %>%
               group_by(continent) %>%
               summarise(mean_lifeExp = mean(lifeExp)),
             aes(xintercept = mean_lifeExp, color = continent),
             linetype = "dashed") +  # Add vertical mean lines
  facet_wrap(~ continent, ncol = 1, scales = "free_y") +  # Stack density plots vertically
  labs(x = "Life Expectancy",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_blank())  # Turn off facet labels


# Combine the box plot and density plot side-by-side
combined_plot <- boxplot | density_plot  # Use patchwork to align the visuals horizontally

# Display the combined plot
print(combined_plot)
  
```

