---
title: "Chi Squared"
author: "Jason Pemberton"
date: "2025-04-01"
output: html_document
---

```{r Load Libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(patchwork)
library(forcats)
```

---
Chi Squared:
1. Goodness of Fit
2. Test of Independence 

Used for categorical data

```{r Load iris data}

head(iris)
```
---

```{r}
flowers <- iris %>% 
  mutate(size = cut(Sepal.Length,
                    breaks = 3,
                    labels = c("Small", "Medium", "Large"))) %>% 
  select(Species, size)

table(flowers)
```

---
H0 (Null Hypothesis): The proportion of flowers that are small, medium and large are equal
H1 (Alt. Hypothesis): The proportion of flowers that are small, medium and large are not equal  
```{r Chi Squared Goodness of Fit Test}
flowers %>% 
  select(size) %>% 
  table() %>% 
  chisq.test()

# same as running chisq.test(table(flowers$size))
# p = 0.000000667
# The probability of the sample data providing a difference of this magnitude
# or more is < 0.05, so we reject the null hypothesis. The proportions are not equal
```

---
H0: The variables are independent, there is no relationship between the variables
H1: The variables are not independent.
Knowing the value of one variable does not help predict the value of another variable
```{r Chi Squared Test of Independence}
# chisq.test(flowers)

flowers %>% 
  table() %>% 
  chisq.test()

my_test <- flowers %>% 
  table() %>% 
  chisq.test()

attributes(my_test)
my_test$statistic
my_test$p.value

# p = 2.2e-16
# We can reject the null hypothesis, they are dependent
```

---

```{r GSS_CAT Goodness of Fit}
my_data <- gss_cat %>% 
  select(marital) %>%
  filter(marital %in% c("Married",
                         "Never married",
                         "Divorced")) %>% 
  mutate(marital = fct_drop(marital))

my_table <- table(my_data)
my_table

# Use your prepared data
my_data <- gss_cat %>% 
  select(marital) %>%
  filter(marital %in% c("Married", "Never married", "Divorced")) %>% 
  mutate(marital = fct_drop(marital))

# Create a column chart
my_data %>%
  count(marital) %>% # Count occurrences of each marital category
  ggplot(aes(x = marital, y = n, fill = marital)) +
  geom_col() +
  labs(
    title = "Count of Marital Status Categories",
    x = "Marital Status",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none") # Remove the legend for simplicity

# H0: The proportions of Married, Never Married, and Divorced is the same in our data
# H1: The proportions are different

# Applying chi squared test to a table
chisq.test(my_table)

# Our p-value is < 0.05, we reject the null hypothesis, they are not equal.
```

