---
title: "GGPLOT - House Price"
author: "Jason Pemberton"
date: "2025-03-10"
output: html_document
---

**VISUALIZATION:**

Creating a basic scatter plot from two continuous variables.

**Step 1:** Load necessary libraries\
- tidyverse: a collection of R packages that help with data science tasks like importing, tidying, manipulating, and visualizing data\
- ggpmisc: defines stats that help annotate the plots based on the results of model fitting

```{r Load Libraries, include=FALSE}
library(tidyverse)
library(ggpmisc)
```

**Step 2:** Load your data into a dataframe then take a look at your data\
- head() returns the first 6 rows of your data (tail, the last 6 rows)\
- summary() returns summary statistics for each variable in your data\
- View() opens your entire data table

```{r Load Data and EDA, echo=FALSE}
df <- read.csv('./data/HousePrices.csv')
head(df)
summary(df)
# View(df)
```

------------------------------------------------------------------------

**Step 3:** Visualize your data using base R commands\
- type the name of your dataframe followed by the \$ symbol to gain access to your variables

```{r Base R Visualization, echo=FALSE, message=FALSE, warning=FALSE}
# Base R scatter plot
plot(df$area, df$price)
hist(df$price)
```

------------------------------------------------------------------------

**Step 4:** Visualizing your data using ggplot

```{r GGPLOT Visualization, echo=FALSE, message=FALSE, warning=FALSE}
# GGPLOT
ggplot(df, aes(x = area, y = price)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE, color="red") +
  stat_poly_eq(aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
               formula = y ~ x, 
               parse = TRUE) +
  labs(title = "Scatterplot of Area vs Price",
       x = "Area (sq ft)",
       y = "Price ($)") +
  theme_minimal()

df %>% 
  ggplot(aes(area, price)) +
  geom_point() +
  geom_smooth(method=lm, se=F) +
  labs(
    title = "House Size vs Price",
    x = "",
    y = "Price (USD)"
  )
```
