---
title: "Visualization - Part 3"
output:
  html_document:
    df_print: paged
---

**Load Libraries**\
Most of what you will need for a simple visualization will be contained in tidyverse.\
Later we will begin searching for additional libraries to carry out more complex machine learning techniques.

```{r Load Libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
```

------------------------------------------------------------------------

The built-in "msleep" dataset contains statistics on the sleep behaviour of a variety of mammals.  

```{r EDA, echo=FALSE, message=FALSE, warning=FALSE}
head(msleep)

summary(msleep)

names(msleep)
```

---

Scatterplot of two continuous variables "body weight" and "brain weight"  
 - Adjust the bodywt filter to see how the plot changes  
 - Add modifiers to the trend line (geom_smooth) to turn on linear regression and remove error bars  
    - method = "lm", se= F   
 
```{r Visualize data with scatterplot, echo=FALSE, message=FALSE, warning=FALSE}
msleep %>% 
  filter(bodywt < 50) %>% 
  ggplot(aes(bodywt, brainwt))+
  geom_point(aes(colour = sleep_total,
                 size = awake))+
  geom_smooth()+
  # geom_text(aes(label = name), vjust = -1, hjust = 0.5) +
  labs(
    x = "Body Weight",
    y = "Brain Weight",
    title = "Brain and Body Weight")+
  theme_minimal()
```

---

Boxplot of one categorical vs one continuous variable "vore" and "brain weight"   
 - Adjust the bodywt filter to see how the plot changes  
 - see what effect coord_flip() has on the visual  

 
```{r Visualize portion of data with boxplot, echo=FALSE, message=FALSE, warning=FALSE}

msleep %>% 
  filter(!is.na(brainwt), !is.na(vore), bodywt < 10) %>% 
  ggplot(aes(x = vore, y = brainwt)) +
  geom_boxplot() +
  theme_minimal() +
  coord_flip() +
  labs(title = "Distribution of Brain Weight by Vore Type",
       x = "Vore",
       y = "Brain Weight (kg)")
```

---

```{r}
# Create histograms faceted by vore type

msleep %>% 
  filter(!is.na(brainwt), !is.na(vore), bodywt < 100) %>%
    ggplot(aes(x = brainwt, fill = vore)) +
      geom_histogram(bins = 10, color = "black", alpha = 0.7) +  # Adjust bins as needed
      facet_wrap(~ vore, ncol = 1, scales = "fixed") +  # One column, free scale per vore
      theme_minimal() +
      labs(title = "Histogram of Brain Weight by Vore Type",
          x = "Brain Weight (kg)",
          y = "Count") +
      scale_fill_brewer(palette = "Set2")  # Nice colour scheme
```


