---
title: "Visualization - Part 5"
author: "Jason Pemberton"
date: "2025-03-11"
output: html_document
---

Load Libraries

```{r Load Libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
```

------------------------------------------------------------------------

Single categorical variable, consider using a bar chart to display the count of occurrences

***fct_intfreq*** is a function from the **forcats** package that reorders the levels of a factor variable based on their frequency, with the most common level appearing first.  

```{r Single Continuous Variable, message=FALSE, warning=FALSE}
msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(fct_infreq(vore))) +
  geom_bar(fill = "light blue") +
  theme_bw()
```

---

Single numeric variable, consider using a histogram    

```{r}
msleep %>% 
  ggplot(aes(awake)) +
  geom_histogram(binwidth = 1, fill = "light green") +
  theme_bw() +
  labs(
    x = "Total Sleep",
    y = NULL,
    title = "Histogram of Total Sleep"
  )
```

---

Single Categorical and Single Numeric Variable  

```{r Option 1: Boxplot, echo=FALSE, message=FALSE, warning=FALSE}
msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(vore, sleep_total)) +
  geom_boxplot() +
  coord_flip() +
  theme_bw()
```

---

```{r Option 2a: Density Plot, echo=FALSE, message=FALSE, warning=FALSE}
msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(sleep_total)) +
  geom_density() +
  facet_wrap(~vore) +
  theme_bw()
```

---

```{r Option 2b: Density Plot Filled by Category, echo=FALSE, message=FALSE, warning=FALSE}
msleep %>% 
  drop_na(vore) %>% 
  # filter(vore == "omni" | vore == "herbi") %>%
  # filter(vore %in% c("omni", "herbi")) %>% 
  ggplot(aes(sleep_total, fill = vore)) +
  geom_density(alpha = 0.4) +
  theme_bw()
```
