---
title: "Visualization"
author: "Jason Pemberton"
date: "2025-03-04"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

**VISUALIZATION:**  
How ggplot works

```{r Load Libraries, include=FALSE}
library(tidyverse)
```

**GGPLOT Components**  

-   data - what data we will be using?  

-   mapping (aesthetics) - what variables will be on each axis?  

-   geometric representation - what type of plot will we be making?  

-   statistics  

-   facet  

-   coordinate space  

-   labels  

-   theme  

```{r R built-in datasets, echo=FALSE}
# What datasets are built in to R?
data()
```

___


```{r Biochemical Oxygen Demand Dataset, echo=FALSE, message=FALSE, warning=FALSE}
# BOD

ggplot(data = BOD,
       mapping = aes(x = Time, y = demand)) +
  geom_point(size=3, colour = "red") +
  geom_line(color="blue")


# ggplot assumes the data is the first argument, aesthetics is second, and your first variable is x, the second is y. 
ggplot(BOD, aes(Time, demand)) +
  geom_point(size=3, colour = "red") +
  geom_line(colour="blue")
```

___

Using the pipe operator %>%, we can pass the data to the ggplot function.  
We no longer have to specify the source data in the ggplot() line. Think  
of %>% as "and then..."  

```{r CO2 Dataset, echo=FALSE, message=FALSE, warning=FALSE}

# CO2
# head(CO2)

# variables are: Plant, Type, Treatment, Concentration, Uptake
# ?CO2


CO2 %>% 
  ggplot(aes(conc, uptake, colour = Treatment)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_smooth(method = lm, se = FALSE) + 
  facet_wrap(~Type) + # use a categorical variable to split the resulting plot into one plot for each category
  labs(
    title = "Concentration of CO2"
  ) +
  theme_bw()
```

___

```{r Line Plots with Points, echo=FALSE}
CO2 %>% 
  ggplot(aes(Treatment, uptake)) +
  geom_boxplot() +
  geom_point(alpha = 0.5, 
             aes(size=conc, # using a continuous variable to control point size
                 colour = Plant)) + # using a categorical variable to control plant colour
  facet_wrap(~Type) +
  coord_flip() + # rotate the visual 90'
  theme_bw() +
  labs(
    title = "Chilled vs Non-Chilled"
  )
```

___

```{r Miles Per Gallon Dataset, echo=FALSE, message=FALSE, warning=FALSE}

# head(mpg)

mpg %>%
  filter(cty < 25) %>% 
  ggplot(aes(displ, cty)) +
  geom_point(aes(colour = drv,
                 size = trans),
             alpha = 0.5) + # alpha controls the point transparency (0-1)
  geom_smooth(method = lm) +
  facet_wrap(~year, nrow = 1) +
  labs(
    x="Engine Size",
    y="Miles per Gallon (mpg in city)",
    title = "Fuel Efficiency"
  ) +
  theme_minimal()
```
