---
title: "DATA 420 - Final Project"
author: "Jason Pemberton"
date: "2023-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(plotly)
library(pastecs)
```

The final project for DATA 420 will give you the opportunity to combine all the skills you have learned throught the course to analyze historical weather patterns.

For this project you will choose a city, obtain the city's latitude and longitude and enter the coordinates into the first .Rmd file. When you knit the file - the script will search the NOAA weather website to locate the nearest weather station to your city. The script will create a data frame that contains historical weather data for your city including: date, temperature, & precipitation (rain/snow).

Part 1:
The script in the first .Rmd will prepare your data for analysis. Your data analysis project should begin with data understanding. Use the appropriate R function to provide you with summary statistics on your data. Also consider visualizing your data with boxplots or histograms to develop a better understanding of what represents "normal" data versus outliers and extremes. 

```{r}

summary(local_weather)

lapply(local_weather, sd)

stat.desc(local_weather)

local_weather %>% 
  ggplot(aes(tmax))+
  geom_histogram()

local_weather %>% 
  ggplot(aes(prcp))+
  geom_histogram()

local_weather %>% 
  ggplot(aes(snow))+
  geom_histogram()
 

boxplot(local_weather$tmax)
boxplot(local_weather$prcp)
boxplot(local_weather$snow)


```

