---
title: "DATA 420 - KNN Credit Risk"
author: "Jason Pemberton"
date: "2023-11-13"
output: 
  html_document:
    toc: true
    toc_float: true
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries, include=FALSE}
# Load the 'class' library for k-NN and corrplot to create a correlation plot of the confusion matrix
library(class)
library(corrplot)
library(ggplot2)
```

# Dataframe

Create a hypothetical financial dataset of 10000 customers with random (normal distribution)
predictor variables "income", "savings", "credit score" are numeric while "risk" (target) must be a factor
```{r Create dataframe}
set.seed(123) # allows you to repeat results when using random number generators
n <- 10000

income <- rnorm(n, mean = 50000, sd = 10000)
savings <- rnorm(n, mean = 20000, sd = 5000)
credit_score <- rnorm(n, mean = 700, sd = 50)
risk <- factor(ifelse(income < 45000 | savings < 15000 | credit_score < 650, "High Risk", "Low Risk"))

# turn these columns into a data frame
financial_data <- data.frame(Income = income, Savings = savings, CreditScore = credit_score, Risk = risk)
```


# Data Partition
Split the dataset into training and testing sets. You cna re-run the model with different 
percentages to test their affect on model accuracy 
```{r}
sample_indices <- sample(1:n, n * 0.7)  # 70% for training
train_data <- financial_data[sample_indices, ]
test_data <- financial_data[-sample_indices, ]
```


# KNN Model
Specify the number of neighbours (k) and the feature columns
```{r knn model}
k <- 100
features <- c("Income", "Savings", "CreditScore") # predictors

# Train the k-NN model
knn_model <- knn(train = train_data[, features], test = test_data[, features], cl = train_data$Risk, k = k)

# Predict on the test data
predictions <- as.factor(knn_model)

# Evaluate the model
confusion_matrix <- table(Actual = test_data$Risk, Predicted = predictions)
print(confusion_matrix)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
cat("Accuracy:", accuracy*100,"%", "\n")

```

# Plot Model
Specify the number of neighbours (k) and the feature columns
```{r plot model}

# Create a data frame for ggplot
confusion_df <- as.data.frame(as.table(confusion_matrix))
names(confusion_df) <- c("Actual", "Predicted", "Count")

# Calculate percentages based on model accuracy
confusion_df$Percentage <- confusion_df$Count / sum(confusion_matrix) * 100

# Plot confusion matrix using ggplot

ggplot(confusion_df, aes(x = Actual, y = Predicted)) +
  geom_tile(aes(fill = Percentage), color = "white") +
  geom_text(aes(label = sprintf("%.2f%%", Percentage)), vjust = 1.5, color = "black") +
  scale_fill_gradient(low = "salmon", high = "green", guide = "none") +
  ggtitle(paste("Confusion Matrix:", "Model Accuracy", accuracy*100, "%")) +
  labs(x = "Actual", y = "Predicted") +
  theme_minimal()
```



