---
title: "DATA420 Monte Carlo Simulation Financial"
author: "Jason Pemberton"
date: "2023-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
library(tidyverse)
```

### **How Monte Carlo Performs Risk Analysis**

Monte Carlo simulation is performed by repeatedly running a model on a simulated outcome based on varying inputs; the inputs are uncertain and variable. A common strategy for modelling uncertainty is to randomly sample values from a probability distribution. This allows you to create thousands of input sets for your model. In this way, you can run thousands of permutations of your model, which has several benefits:

Your output is a large set of results. This means that you have a probability of outcomes rather than simply a single point estimate. Monte Carlo generates a distribution of simulated outcomes. This makes it easy to graph and communicate findings. It is easy to change the assumptions of the models by varying the distribution type or properties of the inputs. You can easily model correlation between input variables.

```{r Sampling Probability Distrubutions}
# sample from an uniform distribution
stats::runif(1, min = 0, max = 1)

# sample from an exponential distribution
stats::rexp(1)

# sample from a normal distribution
stats::rnorm(1)

# sample from a log normal distribution
stats::rlnorm(1)
```

In this example, we create a model of financial asset allocation where you have to decide what portion of wealth should be allocated to risk-free investment or high-risk investment at multiple time periods. In this example, there are two sources of uncertainty:

The uncertain return of the risky asset How much to allocate to each type of investment

The below code snippet shows a simple function that calculates returns based on different asset allocations.

```{r}
calculate_return <- function(alpha) {
  risk_free_rate <- 1.03
  risky_rate <- rnorm(1) * 0.05 + 1
  (1 - alpha) * risk_free_rate + alpha * risky_rate
}
```

-   ***alpha*** is an interaction variable with a range of 1 and 0 that determines how much wealth should be allocated to each asset class in each discrete time period.

-   ***risky_free_rate*** is a fixed yield that doesn't change between periods.

-   ***risky_rate*** is a random continuous variable that is centered on 1.05 to represent uncertainty.

The code below executes 1,000 runs of the model over twelve discrete time periods. When you plot the simulation outputs using ggplot2, you will see a distribution of outcomes. In this case, each line represents the predicted return on investment based on different series of inputs.
```{r}
RUNS <- 1000
DECISION.STEPS <- 12

simulations <- rerun(RUNS, replicate(DECISION.STEPS, runif(1) %>% calculate_return())) %>%
  set_names(paste0("sim", 1:RUNS)) %>%
  map(~ accumulate(., ~ .x * .y)) %>%
  map_dfr(~ tibble(value = .x, step = 1:DECISION.STEPS), .id = "simulation")

simulations %>%
  ggplot(aes(x = step, y = value)) +
  geom_line(aes(color = simulation)) +
  theme(legend.position = "none") +
  ggtitle("Simulations of returns from asset allocation")
```

To make the output data easier to understand, you can summarize the data. For example, you can compute the min, max, and mean of your simulation runs across the time steps. To do this, run the following code:
```{r}
summary_values <- simulations %>%
  group_by(step) %>%
  summarise(mean_return = mean(value), max_return = max(value), min_return = min(value)) %>%
  gather("series", "value", -step)

summary_values %>%
  ggplot(aes(x = step, y = value)) +
  geom_line(aes(color = series)) +
  ggtitle("Mean values from simulations")

```
